---
title: "Lesson 8: Advanced Topics – Mixed Models, Bayesian Analysis, High-Performance Computing, and Shiny"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install required packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("lme4", quietly = TRUE)) install.packages("lme4")
if (!requireNamespace("data.table", quietly = TRUE)) install.packages("data.table")
if (!requireNamespace("shiny", quietly = TRUE)) install.packages("shiny")
if (!requireNamespace("parallel", quietly = TRUE)) install.packages("parallel")

# Load the packages
library(tidyverse)  # For data manipulation and visualization
library(lme4)       # For mixed-effects models
library(data.table) # For high-performance data manipulation
```

# Introduction

Welcome to Lesson 8 of our curriculum for SAS users learning R! In this lesson, we'll explore advanced topics that showcase R's capabilities beyond what we've covered so far. As a SAS user, you're familiar with procedures like PROC MIXED for mixed models or PROC MCMC for Bayesian analysis. R offers similar capabilities through specialized packages, often with more flexibility and extensibility.

This lesson is designed as a survey of advanced capabilities rather than an in-depth tutorial on each topic. The goal is to show you what's possible in R and provide starting points for further exploration.

## Learning Objectives

By the end of this lesson, you will be able to:

1. Understand the basics of mixed-effects models in R
2. Recognize the potential of Bayesian analysis in R
3. Identify strategies for handling large datasets and improving performance
4. Appreciate the capabilities of Shiny for creating interactive web applications
5. Know where to find resources for further learning on these advanced topics

# Mixed-Effects Models

## What Are Mixed-Effects Models?

Mixed-effects models (also called multilevel or hierarchical models) are used when data has a hierarchical or nested structure. Common scenarios include:

- Repeated measurements on the same subjects over time
- Patients nested within hospitals or clinics
- Students nested within classrooms or schools

In SAS, you would use PROC MIXED for linear mixed models or PROC GLIMMIX for generalized linear mixed models. In R, the `lme4` package is the most widely used for fitting mixed-effects models.

## Creating Sample Longitudinal Data

Let's create a sample longitudinal dataset to demonstrate mixed-effects models:

```{r}
# Set seed for reproducibility
set.seed(123)

# Create a longitudinal dataset
n_subjects <- 30
n_timepoints <- 4

# Create the data frame
longitudinal_data <- expand.grid(
  Subject = factor(1:n_subjects),
  Time = factor(1:n_timepoints, labels = c("Baseline", "Month1", "Month3", "Month6"))
)

# Add treatment group (between-subjects factor)
longitudinal_data$Treatment <- factor(rep(c("Placebo", "Drug"), each = n_subjects/2)[longitudinal_data$Subject])

# Add random subject effect (some subjects naturally have higher/lower values)
subject_effect <- rep(rnorm(n_subjects, mean = 0, sd = 10), each = n_timepoints)

# Add fixed effects
time_effect <- c(0, -5, -10, -15)  # Effect of time (everyone improves over time)
treatment_effect <- ifelse(longitudinal_data$Treatment == "Drug", -10, 0)  # Drug lowers values by 10 units

# Generate the outcome variable (e.g., blood pressure)
longitudinal_data$BP <- 140 +                                # Baseline value
  subject_effect +                                          # Random subject effect
  time_effect[as.numeric(longitudinal_data$Time)] +         # Fixed time effect
  treatment_effect +                                        # Fixed treatment effect
  rnorm(nrow(longitudinal_data), mean = 0, sd = 5)          # Random noise

# View the first few rows
head(longitudinal_data)
```

## Visualizing Longitudinal Data

Before fitting a model, let's visualize the data:

```{r}
# Calculate mean BP by time and treatment
summary_data <- longitudinal_data %>%
  group_by(Time, Treatment) %>%
  summarize(
    mean_BP = mean(BP),
    se_BP = sd(BP) / sqrt(n())
  )

# Plot mean BP over time by treatment
ggplot(summary_data, aes(x = Time, y = mean_BP, group = Treatment, color = Treatment)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_BP - se_BP, ymax = mean_BP + se_BP), width = 0.2) +
  labs(title = "Mean Blood Pressure Over Time by Treatment",
       x = "Time",
       y = "Blood Pressure (mmHg)") +
  theme_minimal()

# Plot individual trajectories
ggplot(longitudinal_data, aes(x = Time, y = BP, group = Subject, color = Treatment)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  geom_line(data = summary_data, aes(x = Time, y = mean_BP, group = Treatment), 
            size = 1.5, alpha = 1) +
  labs(title = "Individual Blood Pressure Trajectories",
       x = "Time",
       y = "Blood Pressure (mmHg)") +
  theme_minimal()
```

## Fitting a Mixed-Effects Model

Now, let's fit a mixed-effects model using the `lme4` package:

```{r}
# Fit a linear mixed model with random intercepts for subjects
mixed_model <- lmer(BP ~ Time * Treatment + (1 | Subject), data = longitudinal_data)

# View the model summary
summary(mixed_model)
```

The model formula `BP ~ Time * Treatment + (1 | Subject)` specifies:
- Fixed effects: Time, Treatment, and their interaction
- Random effects: Random intercept for each Subject (the `(1 | Subject)` term)

This is similar to the following SAS PROC MIXED code:

```
proc mixed data=longitudinal_data;
  class Subject Time Treatment;
  model BP = Time Treatment Time*Treatment;
  random intercept / subject=Subject;
run;
```

## Interpreting the Results

The output includes:
- Fixed effects estimates (coefficients for Time, Treatment, and their interaction)
- Random effects variance (how much variation is due to subject differences)
- Residual variance (unexplained variation)

The fixed effects show the estimated impact of time and treatment on blood pressure. The random effects show how much variation exists between subjects.

## Comparing to a Simple Linear Model

Let's compare the mixed model to a simple linear model that ignores the repeated measures structure:

```{r}
# Fit a simple linear model
simple_model <- lm(BP ~ Time * Treatment, data = longitudinal_data)
summary(simple_model)

# Compare AIC (lower is better)
AIC(mixed_model)
AIC(simple_model)
```

The mixed model typically provides a better fit because it accounts for the correlation between measurements from the same subject.

# Bayesian Analysis

## Introduction to Bayesian Analysis

Bayesian analysis is an approach to statistics that uses Bayes' theorem to update probabilities as more information becomes available. Unlike frequentist methods (which provide point estimates and p-values), Bayesian methods provide full probability distributions for parameters.

In SAS, you might use PROC MCMC or PROC BGLIMM for Bayesian analysis. In R, there are several packages for Bayesian modeling, including:

- `rstanarm`: User-friendly interface to Stan for common models
- `brms`: Bayesian regression models using Stan
- `rjags`: Interface to JAGS for Bayesian hierarchical models
- `bayesplot`: Visualization for Bayesian models

For this brief introduction, we'll use the `rstanarm` package if it's installed. If not, we'll just explain the concepts.

```{r, eval=FALSE}
# This code will only run if rstanarm is installed
if (requireNamespace("rstanarm", quietly = TRUE)) {
  library(rstanarm)
  
  # Fit a Bayesian linear model
  bayes_model <- stan_glm(BP ~ Time * Treatment, 
                          data = longitudinal_data,
                          family = gaussian(),
                          prior = normal(0, 10),
                          prior_intercept = normal(140, 20),
                          chains = 2, iter = 1000)
  
  # View the model summary
  print(summary(bayes_model))
  
  # Plot the posterior distributions
  plot(bayes_model, "areas", prob = 0.95)
}
```

## Key Concepts in Bayesian Analysis

Even without running the code, it's important to understand these key Bayesian concepts:

1. **Prior Distributions**: Your beliefs about parameters before seeing the data
2. **Likelihood**: How likely the data is, given the parameters
3. **Posterior Distributions**: Updated beliefs about parameters after seeing the data
4. **Markov Chain Monte Carlo (MCMC)**: Computational method to sample from posterior distributions
5. **Credible Intervals**: Ranges containing a specified probability of the parameter value (unlike confidence intervals)

## Advantages of Bayesian Analysis

Bayesian analysis offers several advantages:
- Incorporates prior knowledge
- Provides full probability distributions for parameters
- Handles small sample sizes better
- Naturally accommodates hierarchical structures
- Allows for more intuitive interpretation of results

## SAS vs R for Bayesian Analysis

SAS has PROC MCMC and PROC BGLIMM for Bayesian analysis, but R's ecosystem for Bayesian modeling is more extensive and actively developed. The Stan ecosystem (accessed through packages like `rstanarm` and `brms`) is particularly powerful and flexible.

# High-Performance Computing

As a SAS user, you're likely familiar with SAS's capabilities for handling large datasets efficiently. R also offers several approaches for improving performance and handling big data.

## The data.table Package

The `data.table` package is a high-performance alternative to data.frame and tibble for data manipulation. It's particularly efficient for large datasets.

```{r}
# Create a larger dataset
set.seed(123)
n <- 1e6  # 1 million rows

# Create a data frame
large_df <- data.frame(
  id = 1:n,
  group = sample(letters[1:5], n, replace = TRUE),
  value = rnorm(n)
)

# Convert to data.table
large_dt <- data.table(large_df)

# Compare performance for a simple aggregation
system.time({
  # dplyr approach
  result_dplyr <- large_df %>%
    group_by(group) %>%
    summarize(mean_value = mean(value))
})

system.time({
  # data.table approach
  result_dt <- large_dt[, .(mean_value = mean(value)), by = group]
})
```

The `data.table` syntax is concise but different from dplyr:
- `DT[i, j, by]` where:
  - `i` is for row filtering (like `filter()`)
  - `j` is for selecting or computing (like `select()` or `summarize()`)
  - `by` is for grouping (like `group_by()`)

## Parallel Processing

R can use multiple cores for parallel processing using the `parallel` package:

```{r}
# Check how many cores are available
num_cores <- parallel::detectCores()
cat("Number of available cores:", num_cores, "\n")

# Example of parallel processing using mclapply
# (works on Mac/Linux; on Windows, use parLapply instead)
system.time({
  # Sequential processing
  result_seq <- lapply(1:10, function(x) {
    Sys.sleep(0.1)  # Simulate work
    return(x^2)
  })
})

if (.Platform$OS.type != "windows") {
  system.time({
    # Parallel processing
    result_par <- parallel::mclapply(1:10, function(x) {
      Sys.sleep(0.1)  # Simulate work
      return(x^2)
    }, mc.cores = min(2, num_cores))
  })
} else {
  cat("On Windows, use parLapply with a cluster instead of mclapply\n")
}
```

## Other High-Performance Strategies

There are several other approaches for improving performance in R:

1. **Use compiled code**: R can interface with C++ through the `Rcpp` package
2. **Memory-efficient data structures**: Packages like `ff` and `bigmemory` for out-of-memory data
3. **Database connections**: Connect directly to databases using `DBI`, `dbplyr`, etc.
4. **Spark integration**: Use `sparklyr` to work with Apache Spark from R
5. **GPU computing**: Packages like `gpuR` for GPU-accelerated computing

## SAS vs R for High-Performance Computing

SAS is known for its efficient handling of large datasets, especially on disk. R traditionally works in-memory, which can be a limitation, but modern R packages and techniques can overcome many of these limitations. The choice between SAS and R for big data often depends on your specific needs and infrastructure.

# Shiny: Interactive Web Applications

One of R's unique features is the ability to create interactive web applications using the `shiny` package. This has no direct equivalent in SAS (though SAS does have products like SAS Visual Analytics).

## What is Shiny?

Shiny is an R package that makes it easy to build interactive web applications directly from R. You don't need to know HTML, CSS, or JavaScript (though they can be helpful for advanced customization).

A Shiny app consists of two main components:
1. **UI (User Interface)**: Defines what the app looks like
2. **Server**: Defines how the app works (the logic)

## A Simple Shiny App

Here's a simple Shiny app that allows users to explore our longitudinal data:

```{r, eval=FALSE}
# This code won't run in the R Markdown document
# It's just to show the structure of a Shiny app

library(shiny)

# Define UI
ui <- fluidPage(
  titlePanel("Blood Pressure Over Time"),
  
  sidebarLayout(
    sidebarPanel(
      selectInput("treatment", "Select Treatment:",
                  choices = c("All", "Placebo", "Drug"),
                  selected = "All")
    ),
    
    mainPanel(
      plotOutput("bpPlot")
    )
  )
)

# Define server logic
server <- function(input, output) {
  
  # Filter data based on user input
  filtered_data <- reactive({
    if (input$treatment == "All") {
      longitudinal_data
    } else {
      filter(longitudinal_data, Treatment == input$treatment)
    }
  })
  
  # Create the plot
  output$bpPlot <- renderPlot({
    ggplot(filtered_data(), aes(x = Time, y = BP, group = Subject, color = Treatment)) +
      geom_line(alpha = 0.3) +
      geom_point(alpha = 0.3) +
      stat_summary(aes(group = Treatment), fun = mean, geom = "line", size = 1.5) +
      labs(title = paste("Blood Pressure Over Time:", input$treatment),
           x = "Time",
           y = "Blood Pressure (mmHg)") +
      theme_minimal()
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```

## Running a Sample Shiny App

While we can't run a Shiny app directly in this R Markdown document, you can run one of the built-in examples:

```{r, eval=FALSE}
# Run a sample Shiny app
shiny::runExample("01_hello")
```

To run this code, you would need to execute it in the R console or create a separate R script.

## Deploying Shiny Apps

Shiny apps can be deployed in several ways:
1. **shinyapps.io**: RStudio's hosting service (free tier available)
2. **Shiny Server**: Open-source or commercial server software
3. **RStudio Connect**: Commercial product for hosting R content

## Use Cases for Shiny in Biomedical Research

Shiny is particularly useful in biomedical research for:
- Interactive data exploration
- Custom dashboards for monitoring clinical trials
- Educational tools for explaining statistical concepts
- Decision support tools for clinicians
- Patient-facing applications for data collection or visualization

# Further Resources

## Mixed-Effects Models
- [lme4 package documentation](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf)
- Book: "Mixed-Effects Models in S and S-PLUS" by Pinheiro and Bates
- [R-bloggers tutorial on mixed models](https://www.r-bloggers.com/2015/10/getting-started-with-mixed-effect-models-in-r/)

## Bayesian Analysis
- [rstanarm vignettes](https://mc-stan.org/rstanarm/articles/)
- [brms package documentation](https://paul-buerkner.github.io/brms/)
- Book: "Statistical Rethinking" by Richard McElreath
- [Bayesian Analysis with R course on Coursera](https://www.coursera.org/learn/bayesian)

## High-Performance Computing
- [data.table documentation](https://rdatatable.gitlab.io/data.table/)
- [Efficient R Programming book](https://csgillespie.github.io/efficientR/)
- [R for Big Data on GitHub](https://github.com/rstudio/bigdataclass)
- [sparklyr documentation](https://spark.rstudio.com/)

## Shiny
- [Shiny tutorial](https://shiny.rstudio.com/tutorial/)
- [Mastering Shiny book](https://mastering-shiny.org/)
- [Shiny Gallery](https://shiny.rstudio.com/gallery/)
- [Shiny Cheat Sheet](https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)

# SAS vs R: Advanced Topics Comparison

Let's summarize the key differences and similarities between SAS and R for these advanced topics:

| Topic | SAS | R |
|-------|-----|---|
| Mixed Models | PROC MIXED, PROC GLIMMIX | lme4, nlme packages |
| Bayesian Analysis | PROC MCMC, PROC BGLIMM | rstanarm, brms, rjags packages |
| High-Performance Computing | SAS High-Performance Analytics, In-database processing | data.table, parallel, Rcpp, sparklyr |
| Interactive Applications | SAS Visual Analytics (commercial) | Shiny (free and open-source) |

Key differences:
1. **Ecosystem**: R has a more diverse and rapidly evolving ecosystem of packages for advanced methods
2. **Cost**: Many advanced capabilities in SAS require additional modules, while R packages are free
3. **Flexibility**: R often offers more flexibility for customization and extension
4. **Learning Curve**: SAS procedures may be easier to use initially, but R packages often offer more options
5. **Integration**: R integrates more easily with other open-source tools and languages

# Exercises

## Exercise 1: Mixed Model Practice

Using the `longitudinal_data` dataset we created, try fitting a mixed-effects model with random intercepts for each subject. After fitting, use `summary()` to see the results. Compare the fixed effect of Time to what you would get if you ran a simple linear model ignoring subject (or what you might get from a repeated measures ANOVA in SAS).

```{r}
# Your code here
```

## Exercise 2: Shiny Exploration

If you have RStudio available, run one of the built-in Shiny examples using `runExample("01_hello")`. Then, outline (in markdown) an idea for a Shiny app you might create for your own data – what inputs and outputs would it have?

*Your response here*

## Exercise 3: Performance Experiment

Compare the performance of base R, dplyr, and data.table for a simple operation like calculating the mean by group. Use `system.time()` to measure the execution time.

```{r}
# Your code here
```

# Conclusion

In this lesson, you've been introduced to several advanced topics in R: mixed-effects models, Bayesian analysis, high-performance computing, and Shiny applications. These capabilities extend R beyond basic statistical analysis and data manipulation, making it a powerful tool for complex biomedical research.

Key takeaways:
- Mixed-effects models in R can handle hierarchical and longitudinal data similar to SAS PROC MIXED
- Bayesian analysis in R offers flexible and powerful tools for probabilistic modeling
- R provides several approaches for improving performance with large datasets
- Shiny allows you to create interactive web applications directly from R

These advanced topics showcase R's flexibility and extensibility. While SAS offers similar capabilities (often through additional modules), R's open-source nature and active community development mean that cutting-edge methods are often available in R first.

In the next and final lesson, we'll explore reproducibility and reporting using R Markdown and Quarto, showing how to create professional, reproducible documents that combine code, results, and narrative.

# Session Information

```{r}
sessionInfo()
``` 