---
title: "Lesson 3: Data Import and Export – Reading and Writing Data (CSV, Excel, SAS)"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Install required packages if not already installed
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
if (!requireNamespace("readxl", quietly = TRUE)) install.packages("readxl")
if (!requireNamespace("haven", quietly = TRUE)) install.packages("haven")
if (!requireNamespace("writexl", quietly = TRUE)) install.packages("writexl")

# Load the packages
library(readr)      # For CSV files
library(readxl)     # For Excel files
library(haven)      # For SAS files
library(writexl)    # For writing Excel files
```

# Introduction

Welcome to Lesson 3 of our curriculum for SAS users learning R! In this lesson, we'll focus on one of the most fundamental tasks in data analysis: importing and exporting data. As a SAS user, you're familiar with PROC IMPORT, INFILE statements, and librefs. In R, we handle these operations differently, but the concepts are similar.

## Learning Objectives

By the end of this lesson, you will be able to:

1. Import data from CSV, Excel, and SAS formats into R
2. Export R data frames to various file formats
3. Understand how R's approach to data storage differs from SAS
4. Use RStudio's import tools and write equivalent code
5. Manage file paths and working directories in R

# Understanding R's Working Directory

In SAS, you use librefs to reference locations where your data is stored. In R, we work with the concept of a "working directory" - the default location where R will look for files and save outputs.

```{r}
# Check your current working directory
getwd()

# You can change it with setwd() if needed
# setwd("C:/Your/Path/Here")  # Uncomment and modify as needed
```

This is somewhat analogous to a SAS libref, but it's a single default location rather than multiple named libraries. You can always specify full paths when needed.

# Importing CSV/Text Data

CSV (Comma-Separated Values) files are one of the most common data formats. In SAS, you might use PROC IMPORT or an INFILE statement to read these. In R, we use functions from the `readr` package (part of the tidyverse).

## Using read_csv()

```{r}
# Create a sample CSV file to work with
write.csv(
  data.frame(
    PatientID = c(101, 102, 103, 104, 105),
    Age = c(45, 52, 34, 60, 41),
    Sex = c("M", "F", "F", "M", "M"),
    Treatment = c("A", "A", "B", "B", "A"),
    BP = c(120, 135, 118, 142, 125)
  ),
  "blood_pressure.csv", 
  row.names = FALSE
)

# Now import it using readr
bp_data <- read_csv("blood_pressure.csv")

# Look at the first few rows
head(bp_data)

# Check the structure
str(bp_data)
```

Notice how `read_csv()` automatically detected the data types. This is similar to how PROC IMPORT works in SAS, but R gives you more control over the import process if needed.

## Import Options

The `read_csv()` function has many options to customize how data is imported:

```{r}
# Import with specific column types
bp_data2 <- read_csv("blood_pressure.csv", 
                     col_types = cols(
                       PatientID = col_integer(),
                       Age = col_integer(),
                       Sex = col_character(),
                       Treatment = col_factor(levels = c("A", "B")),
                       BP = col_double()
                     ))

# Check the structure now
str(bp_data2)
```

This is similar to how you might specify informats in SAS, but with a different syntax.

## Using RStudio's Import Dataset Feature

RStudio also provides a graphical interface for importing data, similar to SAS Enterprise Guide. To use it:

1. Click on "Import Dataset" in the Environment tab
2. Choose the file type (CSV, Excel, etc.)
3. Browse to your file
4. Adjust import options as needed
5. Click "Import"

The tool will also show you the equivalent R code, which you can copy into your script for reproducibility.

# Reading Excel Files

Excel files are common in biomedical research. In SAS, you might use PROC IMPORT with the EXCEL engine or SAS/ACCESS. In R, we use the `readxl` package.

```{r}
# Create a sample Excel file
sample_data <- data.frame(
  PatientID = c(201, 202, 203, 204, 205),
  Glucose = c(95, 110, 105, 95, 120),
  Cholesterol = c(180, 210, 190, 200, 220),
  Visit = c(1, 1, 1, 1, 1)
)

# Write to Excel
write_xlsx(sample_data, "lab_results.xlsx")

# Now read it back
lab_data <- read_excel("lab_results.xlsx")

# Check the data
head(lab_data)
```

## Reading Specific Sheets and Ranges

Excel files often have multiple sheets. You can specify which one to read:

```{r}
# Create a multi-sheet Excel file
sheets <- list("Visit1" = sample_data,
               "Visit2" = transform(sample_data, Visit = 2, Glucose = Glucose + 5))
write_xlsx(sheets, "lab_results_multi.xlsx")

# Read a specific sheet
visit2_data <- read_excel("lab_results_multi.xlsx", sheet = "Visit2")
head(visit2_data)

# You can also read specific cell ranges
# visit1_subset <- read_excel("lab_results_multi.xlsx", range = "A1:C6")
```

This flexibility is similar to what you might achieve with SAS's PROC IMPORT options, but the syntax is different.

# Importing SAS Datasets

For SAS users, being able to read SAS datasets directly is crucial. The `haven` package makes this possible.

```{r, eval=FALSE}
# This code is provided as an example but won't run unless you have a SAS dataset
# Replace "your_sas_data.sas7bdat" with an actual SAS dataset path if available

# Import a SAS dataset
# clinical_data <- read_sas("your_sas_data.sas7bdat")
# head(clinical_data)
```

If you have a SAS dataset available, you can uncomment the code above and replace the filename.

## Handling SAS Labels and Formats

One advantage of `haven` is that it preserves SAS metadata like variable labels and formats:

```{r, eval=FALSE}
# Example of accessing SAS metadata (won't run without a SAS dataset)
# Get variable labels
# sapply(clinical_data, attr, "label")

# Check for SAS formats
# sapply(clinical_data, attr, "format.sas")
```

This helps maintain the context of your data when transitioning from SAS to R.

# Creating Sample SAS-like Data

Since we may not have actual SAS datasets available, let's create a data frame that mimics one, with labels and value labels (similar to SAS formats):

```{r}
# Create a data frame
clinical_data <- data.frame(
  ID = 1:5,
  Sex = c(1, 2, 2, 1, 1),
  Race = c(1, 1, 2, 3, 1),
  Treatment = c(1, 1, 2, 2, 1),
  Response = c(0, 1, 1, 0, 1)
)

# Add labels (similar to SAS LABEL statement)
attr(clinical_data$Sex, "label") <- "Patient Sex"
attr(clinical_data$Race, "label") <- "Patient Race"
attr(clinical_data$Treatment, "label") <- "Treatment Group"
attr(clinical_data$Response, "label") <- "Treatment Response (1=Success)"

# Add value labels (similar to SAS formats)
clinical_data$Sex <- haven::labelled(
  clinical_data$Sex, 
  labels = c("Male" = 1, "Female" = 2)
)

clinical_data$Race <- haven::labelled(
  clinical_data$Race, 
  labels = c("White" = 1, "Black" = 2, "Asian" = 3, "Other" = 4)
)

clinical_data$Treatment <- haven::labelled(
  clinical_data$Treatment, 
  labels = c("Placebo" = 1, "Active" = 2)
)

# View the data
clinical_data

# Check the labels
str(clinical_data)
```

This demonstrates how R can work with labeled data similar to SAS.

# Exporting Data

In SAS, you might use PROC EXPORT or DATA steps to write data to external files. In R, we have similar functions for different file formats.

## Writing to CSV

```{r}
# Export a data frame to CSV
write_csv(bp_data, "bp_data_export.csv")

# With options (like SAS's PROC EXPORT options)
write_csv(bp_data, "bp_data_export_na.csv", na = ".")  # Use . for missing values like SAS
```

## Writing to Excel

```{r}
# Export to Excel
write_xlsx(list(BloodPressure = bp_data, Labs = lab_data), "combined_data.xlsx")
```

This creates an Excel file with multiple sheets, similar to what you might do with multiple PROC EXPORT steps in SAS.

## Writing to SAS Format

While R can't directly write SAS's native .sas7bdat format, the `haven` package provides `write_sas()` which creates SAS transport files (.xpt):

```{r, eval=FALSE}
# Export to SAS transport format
# haven::write_xpt(bp_data, "bp_data.xpt")
```

For most interoperability needs, however, CSV is the simplest format to exchange data between SAS and R.

# SAS vs R: Data Storage Philosophy

Let's discuss some key differences in how SAS and R handle data storage:

1. **Temporary vs. Permanent Storage**:
   - SAS: Distinguishes between WORK (temporary) and permanent libraries
   - R: All objects exist in memory until explicitly saved to disk

2. **Data Persistence**:
   - SAS: Data in permanent libraries persists between sessions
   - R: Data must be explicitly saved (e.g., with `saveRDS()` or `write_csv()`) and loaded in each session

3. **Working with Multiple Datasets**:
   - SAS: References datasets by libref.dataset
   - R: Keeps multiple data frames in memory, referenced by object name

4. **Metadata Handling**:
   - SAS: Rich metadata (labels, formats) integrated into the system
   - R: Metadata can be stored as attributes but requires more explicit handling

# Saving and Loading R Objects

Beyond exporting to external formats, R has its own way to save and load R objects:

```{r}
# Save an R object to an .rds file
saveRDS(bp_data, "bp_data.rds")

# Load it back
bp_data_reloaded <- readRDS("bp_data.rds")

# Save multiple objects to an .RData file
save(bp_data, lab_data, file = "multiple_datasets.RData")

# Load them back (objects will have the same names)
# load("multiple_datasets.RData")
```

This is somewhat similar to SAS permanent datasets, but the approach is different.

# Exercises

## Exercise 1: Importing CSV Data

Using `read_csv()`, import the "blood_pressure.csv" file we created earlier. Check the first few rows with `head()` and the structure with `str()`. In a text response, note how the data types were interpreted (e.g., are treatments read as character or factor?).

```{r}
# Your code here
```

*Your response here*

## Exercise 2: Importing Excel Data

Import the "lab_results.xlsx" file using `read_excel()`. Verify that the data has been read correctly by summarizing a few columns (e.g., use `summary()` or `table()` on a column).

```{r}
# Your code here
```

## Exercise 3: Working with Labeled Data and Exporting

Take the `clinical_data` data frame we created (with labels and value labels) and export it as a CSV file. Then, import it back and check if the labels were preserved. What happened to the labels? How might you preserve this metadata when moving data between R sessions?

```{r}
# Your code here
```

*Your response here*

# Conclusion

In this lesson, you've learned how to import and export data in R from various formats, with special attention to SAS datasets. You've seen that while the syntax differs from SAS, the concepts are similar, and R provides flexible tools for data interchange.

Key takeaways:
- R uses functions like `read_csv()`, `read_excel()`, and `read_sas()` instead of PROC IMPORT
- Data in R lives in memory until explicitly saved, unlike SAS's permanent libraries
- R can preserve SAS metadata like labels and formats using the `haven` package
- For sharing data between SAS and R, CSV is often the simplest approach

In the next lesson, we'll explore data wrangling with the tidyverse, which provides powerful tools for transforming and combining data, similar to SAS DATA steps and PROC SQL.

# Session Information

```{r}
sessionInfo()
``` 