---
title: "Lesson 5: Exploratory Data Analysis (EDA) – Summarization and Visualization"
author: "For SAS Users Learning R"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Install required packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")

# Load the packages
library(tidyverse)  # Loads ggplot2, dplyr, and other tidyverse packages
```

# Introduction

Welcome to Lesson 5 of our curriculum for SAS users learning R! In this lesson, we'll explore techniques for Exploratory Data Analysis (EDA) in R. As a SAS user, you're familiar with procedures like PROC MEANS, PROC FREQ, and PROC SGPLOT for summarizing and visualizing data. R offers similar capabilities with more flexibility and interactivity.

## Learning Objectives

By the end of this lesson, you will be able to:

1. Generate summary statistics for continuous and categorical variables
2. Detect and handle missing values in your data
3. Create basic visualizations using ggplot2
4. Customize plots for publication quality
5. Interpret exploratory analyses in a biomedical context
6. Understand the differences between SAS and R approaches to EDA

# Creating Sample Data

Let's create a sample clinical trial dataset to work with throughout this lesson:

```{r}
# Set seed for reproducibility
set.seed(123)

# Create a clinical trial dataset
clinical_trial <- data.frame(
  PatientID = 1:100,
  Age = round(rnorm(100, mean = 55, sd = 10)),
  Sex = factor(sample(c("Male", "Female"), 100, replace = TRUE)),
  Treatment = factor(sample(c("Placebo", "Drug A", "Drug B"), 100, replace = TRUE)),
  Cholesterol = round(rnorm(100, mean = 200, sd = 25)),
  BMI = round(rnorm(100, mean = 26, sd = 4), 1),
  Outcome = factor(sample(c("Improved", "No Change", "Worsened"), 100, replace = TRUE, 
                          prob = c(0.5, 0.3, 0.2)))
)

# Introduce some missing values
clinical_trial$Cholesterol[sample(1:100, 5)] <- NA
clinical_trial$BMI[sample(1:100, 8)] <- NA

# View the first few rows
head(clinical_trial)
```

# Summarizing Data

## Basic Summary Statistics

In SAS, you might use PROC MEANS or PROC UNIVARIATE to get summary statistics. In R, there are several ways to do this:

```{r}
# Basic summary of the entire dataset
summary(clinical_trial)

# Summary of a specific variable
summary(clinical_trial$Cholesterol)

# More detailed statistics for a numeric variable
cholesterol_stats <- clinical_trial %>%
  summarize(
    n = n(),
    n_missing = sum(is.na(Cholesterol)),
    mean = mean(Cholesterol, na.rm = TRUE),
    median = median(Cholesterol, na.rm = TRUE),
    sd = sd(Cholesterol, na.rm = TRUE),
    min = min(Cholesterol, na.rm = TRUE),
    q1 = quantile(Cholesterol, 0.25, na.rm = TRUE),
    q3 = quantile(Cholesterol, 0.75, na.rm = TRUE),
    max = max(Cholesterol, na.rm = TRUE)
  )

cholesterol_stats
```
> **Summarize it!** Run the chunks above. Compare the output of the basic `summary()` function with the more detailed statistics we calculated using `summarize()`. Notice how `na.rm = TRUE` is essential for calculations involving `Cholesterol` because it has missing values.


## Summarizing by Group

Similar to PROC MEANS with a CLASS statement in SAS, we can use `group_by()` and `summarize()` to get statistics by group:

```{r}
# Summary statistics by treatment group
clinical_trial %>%
  group_by(Treatment) %>%
  summarize(
    n = n(),
    mean_chol = mean(Cholesterol, na.rm = TRUE),
    sd_chol = sd(Cholesterol, na.rm = TRUE),
    mean_bmi = mean(BMI, na.rm = TRUE),
    sd_bmi = sd(BMI, na.rm = TRUE)
  )

# Summary statistics by treatment and sex
clinical_trial %>%
  group_by(Treatment, Sex) %>%
  summarize(
    n = n(),
    mean_chol = mean(Cholesterol, na.rm = TRUE),
    sd_chol = sd(Cholesterol, na.rm = TRUE)
  )
```
> **Group Summaries:** Run these chunks. See how `group_by()` allows `summarize()` to calculate statistics for each treatment group, or even combinations like Treatment and Sex?


## Frequency Tables

In SAS, you might use PROC FREQ to create frequency tables. In R, we can use `table()` or dplyr's `count()`:

```{r}
# Simple frequency table
table(clinical_trial$Treatment)

# Proportions
prop.table(table(clinical_trial$Treatment))

# Cross-tabulation
treatment_outcome_table <- table(clinical_trial$Treatment, clinical_trial$Outcome)
treatment_outcome_table

# Proportions within each treatment
prop.table(treatment_outcome_table, 1)  # Row proportions

# Using dplyr's count
clinical_trial %>%
  count(Treatment, Outcome) %>%
  pivot_wider(names_from = Outcome, values_from = n, values_fill = 0)
```
> **Frequencies:** Run these chunks. Compare the output of base R's `table()` and `prop.table()` with `dplyr`'s `count()` followed by `pivot_wider()`. Both achieve similar goals for creating frequency tables.


# Handling Missing Values in EDA

In SAS, missing values are represented by a period (.) for numeric variables and blank for character variables. In R, missing values are represented by `NA` for all types of variables.

## Detecting Missing Values

```{r}
# Check for missing values in the entire dataset
colSums(is.na(clinical_trial))

# Visualize missing values pattern
clinical_trial %>%
  summarize(across(everything(), ~ sum(is.na(.)))) %>%
> **Find the Missing!** Run the `colSums(is.na(...))` command to get a quick count of NAs per column. Then, run the `ggplot` chunk. This plot gives a nice visual overview of where the missingness is concentrated.

  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  ggplot(aes(x = reorder(Variable, Missing_Count), y = Missing_Count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Missing Values by Variable", x = "Variable", y = "Count of Missing Values")
```

## Impact of Missing Values on Analysis

Let's see how missing values affect our summary statistics:

```{r}
# Mean with and without handling NAs
mean(clinical_trial$Cholesterol)  # Returns NA if any value is NA
> **NA Impact:** Run these lines. See how `mean()` returns `NA` by default if any input is `NA`? Using `na.rm = TRUE` is crucial. Also, check how many rows are lost if we only keep `complete.cases()`.

mean(clinical_trial$Cholesterol, na.rm = TRUE)  # Removes NAs before calculation

# Complete cases only
complete_cases <- clinical_trial[complete.cases(clinical_trial), ]
cat("Complete cases:", nrow(complete_cases), "out of", nrow(clinical_trial), "\n")
```

## Simple Imputation

While advanced imputation methods are beyond the scope of this lesson, here's a simple example of imputing missing values with the mean:

```{r}
# Impute missing Cholesterol values with the mean
clinical_trial_imputed <- clinical_trial %>%
  mutate(Cholesterol_Imputed = ifelse(is.na(Cholesterol), 
                                     mean(Cholesterol, na.rm = TRUE), 
> **Simple Fix:** Run this imputation example. Compare the `summary()` outputs. Notice how the mean hasn't changed (as expected with mean imputation), but the `NA's` count is now 0 for the imputed variable. (Remember, mean imputation is simple but often not the best approach!)

                                     Cholesterol))

# Compare original and imputed
summary(clinical_trial$Cholesterol)
summary(clinical_trial_imputed$Cholesterol_Imputed)
```

# Introduction to ggplot2

In SAS, you might use PROC SGPLOT or PROC GPLOT for visualization. In R, the ggplot2 package (part of the tidyverse) is the most popular plotting system. It's based on the "Grammar of Graphics," which provides a structured approach to creating visualizations.

> **Your First Plot!** Run this chunk. Your scatter plot should appear in the Plots pane in RStudio (usually bottom-right). This is the basic structure: data, aesthetics, geometry.

## Basic Structure of a ggplot

A ggplot consists of three main components:
1. Data: The dataset to visualize
2. Aesthetics (aes): Mapping variables to visual properties
3. Geometries (geom): The type of plot to create

```{r}
# Basic scatter plot
ggplot(clinical_trial, aes(x = Age, y = Cholesterol)) +
  geom_point()
```

This is similar to a PROC SGPLOT SCATTER statement in SAS, but the syntax is different.

## Adding Layers
> **Layer Up!** Run this chunk. See how we added a `geom_smooth()` layer for the trend line and `labs()` to add titles and labels? ggplot builds plots layer by layer.


One of the strengths of ggplot2 is the ability to add layers to a plot:

```{r}
# Scatter plot with a trend line
ggplot(clinical_trial, aes(x = Age, y = Cholesterol)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Cholesterol vs. Age",
       x = "Age (years)",
       y = "Cholesterol (mg/dL)")
```

> **Add Some Color!** Run this chunk. By putting `color = Treatment` inside `aes()`, we told ggplot to map the Treatment variable to the color aesthetic.

## Mapping Aesthetics to Variables

We can map other variables to visual properties like color, shape, or size:

```{r}
# Color points by Treatment
ggplot(clinical_trial, aes(x = Age, y = Cholesterol, color = Treatment)) +
  geom_point() +
  labs(title = "Cholesterol vs. Age by Treatment Group",
       x = "Age (years)",
       y = "Cholesterol (mg/dL)")
```

# Common Plot Types

## Histogram

Similar to PROC UNIVARIATE or PROC SGPLOT HISTOGRAM in SAS:

```{r}
# Basic histogram
ggplot(clinical_trial, aes(x = Cholesterol)) +
> **Histograms & Density:** Run these chunks. Histograms show counts in bins, while density plots show a smoothed distribution. Both are great for understanding the shape of your data.

  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  labs(title = "Distribution of Cholesterol",
       x = "Cholesterol (mg/dL)",
       y = "Count")

# Density plot
ggplot(clinical_trial, aes(x = Cholesterol)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(title = "Density of Cholesterol",
       x = "Cholesterol (mg/dL)",
       y = "Density")
```

## Boxplot

Similar to PROC SGPLOT VBOX in SAS:

```{r}
# Boxplot by treatment
ggplot(clinical_trial, aes(x = Treatment, y = Cholesterol, fill = Treatment)) +
  geom_boxplot() +
> **Boxplots:** Run these chunks. Boxplots are excellent for comparing distributions across groups. Adding `geom_jitter()` shows the individual data points too!

  labs(title = "Cholesterol by Treatment Group",
       x = "Treatment",
       y = "Cholesterol (mg/dL)")

# Add individual points
ggplot(clinical_trial, aes(x = Treatment, y = Cholesterol, fill = Treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(title = "Cholesterol by Treatment Group",
       x = "Treatment",
       y = "Cholesterol (mg/dL)")
```

## Bar Chart

Similar to PROC SGPLOT VBAR in SAS:

```{r}
# Bar chart of outcome counts
ggplot(clinical_trial, aes(x = Outcome, fill = Outcome)) +
  geom_bar() +
  labs(title = "Distribution of Outcomes",
       x = "Outcome",
       y = "Count")

# Stacked bar chart
ggplot(clinical_trial, aes(x = Treatment, fill = Outcome)) +
> **Bar Charts:** Run these chunks. `geom_bar()` is used for counts of categorical variables. Notice how `position = "stack"` or `position = "fill"` changes how the bars are displayed when using `fill`.

  geom_bar(position = "stack") +
  labs(title = "Outcomes by Treatment Group",
       x = "Treatment",
       y = "Count")

# Proportional stacked bar chart
ggplot(clinical_trial, aes(x = Treatment, fill = Outcome)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Outcomes by Treatment Group",
       x = "Treatment",
       y = "Proportion")
```

## Line Plot

For longitudinal data, we often use line plots. Let's create a simulated longitudinal dataset:

```{r}
# Create a longitudinal dataset
set.seed(456)
longitudinal_data <- expand.grid(
  PatientID = 1:20,
  Visit = c("Baseline", "Month1", "Month3", "Month6")
)

longitudinal_data$Measurement <- 100 + 
  rep(rnorm(20, mean = 0, sd = 10), each = 4) +  # Patient-specific baseline
  ifelse(longitudinal_data$Visit == "Baseline", 0,
         ifelse(longitudinal_data$Visit == "Month1", -5,
                ifelse(longitudinal_data$Visit == "Month3", -10, -15))) +  # Time effect
  rnorm(nrow(longitudinal_data), mean = 0, sd = 5)  # Random noise

# Add a treatment group
longitudinal_data$Treatment <- factor(rep(rep(c("Placebo", "Active"), each = 10), each = 4))

# Convert Visit to a factor with the correct order
longitudinal_data$Visit <- factor(longitudinal_data$Visit, 
                                 levels = c("Baseline", "Month1", "Month3", "Month6"))

# Calculate mean by visit and treatment
longitudinal_summary <- longitudinal_data %>%
  group_by(Visit, Treatment) %>%
  summarize(
    mean_measurement = mean(Measurement),
    se_measurement = sd(Measurement) / sqrt(n())
  )

> **Line Plots:** Run this chunk (it includes code to create sample longitudinal data first). Line plots are essential for visualizing trends over time. Notice how we calculated summary stats first before plotting the means and error bars.

# Line plot
ggplot(longitudinal_summary, aes(x = Visit, y = mean_measurement, 
                                group = Treatment, color = Treatment)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_measurement - se_measurement, 
                    ymax = mean_measurement + se_measurement), 
                width = 0.2) +
  labs(title = "Mean Measurement Over Time by Treatment Group",
       x = "Visit",
       y = "Mean Measurement")
```

# Enhancing Plots

## Themes

ggplot2 provides several built-in themes to change the overall appearance of a plot:

```{r}
> **Themes:** Run these chunks (after running the initial plot `p`). Themes are a quick way to change the overall look of your plot.

# Default theme
p <- ggplot(clinical_trial, aes(x = Treatment, y = Cholesterol, fill = Treatment)) +
  geom_boxplot() +
  labs(title = "Cholesterol by Treatment Group",
       x = "Treatment",
       y = "Cholesterol (mg/dL)")

# Apply different themes
p + theme_minimal()
p + theme_classic()
p + theme_bw()
```
> **Custom Colors:** Run this chunk. `scale_fill_manual()` gives you precise control over the colors used.


## Customizing Colors

```{r}
# Custom colors
ggplot(clinical_trial, aes(x = Treatment, fill = Treatment)) +
  geom_bar() +
  scale_fill_manual(values = c("Placebo" = "gray", "Drug A" = "skyblue", "Drug B" = "darkblue")) +
  labs(title = "Count by Treatment Group",
       x = "Treatment",
       y = "Count")
```

## Faceting

> **Faceting:** Run this chunk. `facet_wrap()` creates subplots for each level of the specified variable (Sex, in this case). It's great for comparing patterns across groups.

Faceting allows you to create multiple plots based on a categorical variable:

```{r}
# Facet by Sex
ggplot(clinical_trial, aes(x = Age, y = Cholesterol, color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Sex) +
  labs(title = "Cholesterol vs. Age by Treatment and Sex",
       x = "Age (years)",
       y = "Cholesterol (mg/dL)")
```

# SAS vs R: Visualization Comparison

Let's compare how you might create a similar plot in SAS versus R:

**SAS (PROC SGPLOT)**:
```
proc sgplot data=clinical_trial;
  vbox Cholesterol / category=Treatment group=Sex;
  title "Cholesterol by Treatment and Sex";
  xaxis label="Treatment Group";
> **Plot Comparison:** Run this ggplot chunk. Compare its structure to the SAS code above it. Both create boxplots grouped by Treatment and Sex, but the syntax is quite different.

  yaxis label="Cholesterol (mg/dL)";
run;
```

**R (ggplot2)**:
```{r}
ggplot(clinical_trial, aes(x = Treatment, y = Cholesterol, fill = Sex)) +
  geom_boxplot() +
  labs(title = "Cholesterol by Treatment and Sex",
       x = "Treatment Group",
       y = "Cholesterol (mg/dL)")
```

Key differences:
1. **Syntax**: SAS uses a procedural approach with statements, while ggplot2 uses a layered approach with the `+` operator.
2. **Flexibility**: ggplot2 offers more customization options and is more extensible.
3. **Output**: SAS procedures typically produce static output, while ggplot2 objects can be modified, saved, and reused.
4. **Integration**: In R Markdown, plots appear directly in the document, making it easier to combine analysis and visualization.

# Interpreting Results in a Biomedical Context

Let's interpret some of our exploratory analyses in the context of a clinical trial:

```{r}
# Summary statistics by treatment
treatment_summary <- clinical_trial %>%
  group_by(Treatment) %>%
  summarize(
    n = n(),
    mean_chol = mean(Cholesterol, na.rm = TRUE),
    sd_chol = sd(Cholesterol, na.rm = TRUE),
    mean_age = mean(Age),
    sd_age = sd(Age)
> **Interpret:** Run these summary chunks again. Looking at the numbers and the plots together helps build a story about the data.

  )

treatment_summary

# Outcome by treatment
outcome_by_treatment <- clinical_trial %>%
  count(Treatment, Outcome) %>%
  group_by(Treatment) %>%
  mutate(Proportion = n / sum(n))

outcome_by_treatment
```

**Interpretation**:
- The mean cholesterol levels appear to be lower in the Drug B group compared to Placebo and Drug A.
- The age distributions are similar across treatment groups, suggesting good randomization.
- The proportion of patients with "Improved" outcomes is higher in the Drug B group compared to Placebo.

These observations would guide further formal statistical testing, which we'll cover in Lesson 7.

# Exercises

## Exercise 1: Summary Statistics

Using the `clinical_trial` dataset, calculate the following. **Type your code in the chunk below and run it.**
1. The overall mean and median of Cholesterol
2. The mean Cholesterol by Treatment group
3. A frequency table of Outcome by Treatment

Interpret these results in a short text: do the groups differ noticeably in Cholesterol or outcome counts?

```{r}
# Your code here
```

*Your interpretation here*

## Exercise 2: Creating Visualizations

1. Create a histogram of a numeric variable (e.g., BMI) using ggplot2. Customize the bins if necessary and add a title. **Type your code in the chunk below and run it.**
2. Create a boxplot of that variable by a categorical grouping (e.g., BMI by Treatment).
3. Ensure your plots have proper axis labels and interpret what the plots show.

```{r}
# Your code here
```

*Your interpretation here*

## Exercise 3: Longitudinal Data Visualization

Using the `longitudinal_data` dataset, create a line plot showing the mean Measurement over time for each Treatment group. Add error bars representing the standard error of the mean. **Type your code in the chunk below and run it.**

```{r}
# Your code here
```

*Your interpretation here*

# Conclusion

In this lesson, you've learned how to perform exploratory data analysis in R, including summarizing data, handling missing values, and creating visualizations with ggplot2. You've seen how these techniques compare to similar approaches in SAS, and how to interpret the results in a biomedical context.

Key takeaways:
- R provides flexible tools for summarizing data, similar to SAS's PROC MEANS and PROC FREQ
- Missing values (NA) require explicit handling in R, often using the na.rm=TRUE argument
- ggplot2 offers a powerful grammar for creating customizable visualizations
- The layered approach of ggplot2 differs from SAS's procedural approach but offers more flexibility
- Exploratory analysis in R can be interactive and iterative, allowing for quick insights

In the next lesson, we'll explore how to write custom functions and control structures in R, which will help you automate repetitive tasks and create reusable code.

# Session Information

```{r}
sessionInfo()
``` 