---
title: "Lesson 7: Statistical Analysis in R – from Descriptive Stats to Modeling"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install required packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("survival", quietly = TRUE)) install.packages("survival")
if (!requireNamespace("survminer", quietly = TRUE)) install.packages("survminer")
if (!requireNamespace("broom", quietly = TRUE)) install.packages("broom")

# Load the packages
library(tidyverse)  # For data manipulation and visualization
library(survival)   # For survival analysis
library(survminer)  # For survival plots
library(broom)      # For tidying model outputs
```

# Introduction

Welcome to Lesson 7 of our curriculum for SAS users learning R! In this lesson, we'll explore statistical analysis in R, from basic descriptive statistics to more complex modeling techniques. As a SAS user, you're familiar with procedures like PROC TTEST, PROC GLM, and PROC LOGISTIC. R offers similar capabilities through functions, with a more flexible and interactive approach.

## Learning Objectives

By the end of this lesson, you will be able to:

1. Perform basic descriptive statistics and hypothesis tests in R
2. Conduct ANOVA and fit linear regression models
3. Implement logistic regression for binary outcomes
4. Perform basic survival analysis
5. Interpret statistical results in R
6. Understand the parallels between SAS procedures and R functions

# Creating Sample Data

Let's create a sample clinical trial dataset to work with throughout this lesson:

```{r}
# Set seed for reproducibility
set.seed(123)

# Create a clinical trial dataset
clinical_trial <- data.frame(
  PatientID = 1:100,
  Age = round(rnorm(100, mean = 55, sd = 10)),
  Sex = factor(sample(c("Male", "Female"), 100, replace = TRUE)),
  Treatment = factor(sample(c("Placebo", "Drug A", "Drug B"), 100, replace = TRUE)),
  Baseline_BP = round(rnorm(100, mean = 140, sd = 15)),
  Final_BP = NA,  # We'll calculate this based on treatment effect
  Cholesterol = round(rnorm(100, mean = 200, sd = 25)),
  Outcome = factor(NA, levels = c("Failure", "Success"))  # Binary outcome
)

# Add treatment effects
clinical_trial$Final_BP <- clinical_trial$Baseline_BP + 
  ifelse(clinical_trial$Treatment == "Placebo", rnorm(100, -5, 10),
         ifelse(clinical_trial$Treatment == "Drug A", rnorm(100, -15, 10),
                rnorm(100, -25, 10)))  # Drug B has strongest effect

# Create binary outcome based on BP reduction
clinical_trial$Outcome <- factor(
  ifelse(clinical_trial$Final_BP < clinical_trial$Baseline_BP - 10, "Success", "Failure"),
  levels = c("Failure", "Success")
)

# Calculate BP change
clinical_trial$BP_Change <- clinical_trial$Final_BP - clinical_trial$Baseline_BP

# View the first few rows
head(clinical_trial)
```

# Descriptive Statistics and Hypothesis Tests

## Descriptive Statistics

In SAS, you might use PROC MEANS or PROC UNIVARIATE for descriptive statistics. In R, there are several approaches:

```{r}
# Basic summary statistics
summary(clinical_trial[, c("Age", "Baseline_BP", "Final_BP", "BP_Change")])

# Group-wise statistics using dplyr
clinical_trial %>%
  group_by(Treatment) %>%
  summarize(
    n = n(),
    mean_age = mean(Age),
    mean_baseline_bp = mean(Baseline_BP),
    mean_final_bp = mean(Final_BP),
    mean_bp_change = mean(BP_Change),
    sd_bp_change = sd(BP_Change)
  )
```

## Two-Sample t-test

In SAS, you would use PROC TTEST. In R, we use the `t.test()` function:

```{r}
# Compare BP change between Placebo and Drug B
placebo_data <- clinical_trial %>% filter(Treatment == "Placebo")
drugB_data <- clinical_trial %>% filter(Treatment == "Drug B")

t_test_result <- t.test(placebo_data$BP_Change, drugB_data$BP_Change)
t_test_result
```

Let's compare this to what you would see in SAS:

**SAS PROC TTEST:**
```
proc ttest data=clinical_trial;
  class Treatment;
  var BP_Change;
  where Treatment in ("Placebo", "Drug B");
run;
```

The R output provides:
- The t-statistic
- Degrees of freedom
- p-value
- Confidence interval for the difference in means
- Sample means

This is similar to what PROC TTEST provides, though the format is different.

## Chi-Square Test

For categorical data, SAS uses PROC FREQ with the CHISQ option. In R, we use `chisq.test()`:

```{r}
# Create a contingency table
outcome_table <- table(clinical_trial$Treatment, clinical_trial$Outcome)
outcome_table

# Perform chi-square test
chi_sq_result <- chisq.test(outcome_table)
chi_sq_result
```

Let's visualize this relationship:

```{r}
# Bar plot of outcomes by treatment
ggplot(clinical_trial, aes(x = Treatment, fill = Outcome)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Outcomes by Treatment",
       y = "Proportion",
       x = "Treatment Group") +
  theme_minimal()
```

# ANOVA and Linear Models

## One-Way ANOVA

In SAS, you would use PROC GLM or PROC ANOVA. In R, we can use `aov()` or `lm()`:

```{r}
# One-way ANOVA using aov()
anova_result <- aov(BP_Change ~ Treatment, data = clinical_trial)
summary(anova_result)

# Same analysis using lm()
lm_anova <- lm(BP_Change ~ Treatment, data = clinical_trial)
anova(lm_anova)
```

The output shows:
- Degrees of freedom
- Sum of squares
- Mean squares
- F-value
- p-value

This is similar to the ANOVA table from PROC GLM in SAS.

## Post-hoc Tests

For pairwise comparisons after ANOVA, SAS might use the LSMEANS statement with adjustments. In R, we can use `TukeyHSD()`:

```{r}
# Tukey's Honest Significant Difference
tukey_result <- TukeyHSD(anova_result)
tukey_result
```

## Linear Regression

In SAS, you would use PROC REG or PROC GLM. In R, we use `lm()`:

```{r}
# Simple linear regression
lm_simple <- lm(Final_BP ~ Baseline_BP, data = clinical_trial)
summary(lm_simple)

# Multiple linear regression
lm_multiple <- lm(Final_BP ~ Baseline_BP + Age + Treatment, data = clinical_trial)
summary(lm_multiple)
```

The output includes:
- Coefficients with standard errors, t-values, and p-values
- R-squared and adjusted R-squared
- F-statistic and overall p-value

This is similar to the output from PROC REG in SAS, though the format is different.

## Visualizing the Model

Let's visualize the relationship between baseline and final BP, colored by treatment:

```{r}
ggplot(clinical_trial, aes(x = Baseline_BP, y = Final_BP, color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Relationship between Baseline and Final BP",
       x = "Baseline BP (mmHg)",
       y = "Final BP (mmHg)") +
  theme_minimal()
```

## Extracting Model Information

In SAS, you might use ODS OUTPUT to capture model results. In R, we can use the `broom` package to tidy model outputs:

```{r}
# Get tidy coefficient table
tidy(lm_multiple)

# Get model performance metrics
glance(lm_multiple)

# Get fitted values and residuals
augment(lm_multiple) %>% head()
```

## Diagnostic Plots

In SAS, you might use the PLOTS option in PROC REG. In R, we can use the built-in `plot()` function or create custom plots:

```{r}
# Basic diagnostic plots
par(mfrow = c(2, 2))
plot(lm_multiple)
```

# Logistic Regression

In SAS, you would use PROC LOGISTIC. In R, we use `glm()` with `family = binomial`:

```{r}
# Logistic regression
logistic_model <- glm(Outcome ~ Treatment + Age + Baseline_BP, 
                      family = binomial(link = "logit"), 
                      data = clinical_trial)
summary(logistic_model)
```

The output includes:
- Coefficients with standard errors, z-values, and p-values
- AIC (Akaike Information Criterion)
- Null and residual deviance

This is similar to the output from PROC LOGISTIC in SAS, though the format is different.

## Odds Ratios

In SAS, PROC LOGISTIC provides odds ratios directly. In R, we need to exponentiate the coefficients:

```{r}
# Calculate odds ratios and confidence intervals
exp(cbind(OR = coef(logistic_model), confint(logistic_model)))
```

## Predicted Probabilities

Let's calculate and visualize predicted probabilities:

```{r}
# Add predicted probabilities to the dataset
clinical_trial$predicted_prob <- predict(logistic_model, type = "response")

# Plot predicted probabilities by treatment
ggplot(clinical_trial, aes(x = Baseline_BP, y = predicted_prob, color = Treatment)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Predicted Probability of Success by Baseline BP",
       x = "Baseline BP (mmHg)",
       y = "Predicted Probability of Success") +
  theme_minimal()
```

# Survival Analysis

Survival analysis is common in biomedical research. In SAS, you would use PROC LIFETEST and PROC PHREG. In R, we use the `survival` package.

Let's create a survival dataset:

```{r}
# Create a survival dataset
set.seed(456)
n <- 200
survival_data <- data.frame(
  PatientID = 1:n,
  Treatment = factor(sample(c("Placebo", "Drug A", "Drug B"), n, replace = TRUE)),
  Age = round(rnorm(n, mean = 60, sd = 12)),
  Time = NA,
  Status = NA
)

# Generate survival times based on treatment
# Placebo: baseline hazard
# Drug A: 20% reduction in hazard
# Drug B: 40% reduction in hazard
survival_data$Time <- round(rexp(n, rate = ifelse(survival_data$Treatment == "Placebo", 1/20,
                                           ifelse(survival_data$Treatment == "Drug A", 1/25, 1/33))))

# Add censoring (about 30% of patients)
survival_data$Status <- sample(c(1, 0), n, replace = TRUE, prob = c(0.7, 0.3))

# View the first few rows
head(survival_data)
```

## Kaplan-Meier Curves

In SAS, you would use PROC LIFETEST. In R, we use `survfit()`:

```{r}
# Create a survival object
surv_obj <- Surv(time = survival_data$Time, event = survival_data$Status)

# Fit Kaplan-Meier curves by treatment
km_fit <- survfit(surv_obj ~ Treatment, data = survival_data)

# Summary of the Kaplan-Meier fit
summary(km_fit)

# Plot Kaplan-Meier curves
ggsurvplot(km_fit, data = survival_data, 
           pval = TRUE,  # Add p-value from log-rank test
           risk.table = TRUE,  # Add risk table
           conf.int = TRUE,  # Add confidence intervals
           xlab = "Time (months)",
           ylab = "Survival Probability",
           title = "Kaplan-Meier Curves by Treatment Group")
```

## Cox Proportional Hazards Model

In SAS, you would use PROC PHREG. In R, we use `coxph()`:

```{r}
# Fit Cox proportional hazards model
cox_model <- coxph(surv_obj ~ Treatment + Age, data = survival_data)
summary(cox_model)
```

The output includes:
- Coefficients with standard errors, z-values, and p-values
- Hazard ratios (exp(coef)) with confidence intervals
- Concordance, likelihood ratio test, Wald test, and score (logrank) test

This is similar to the output from PROC PHREG in SAS.

## Hazard Ratios

Let's extract and visualize the hazard ratios:

```{r}
# Extract hazard ratios and confidence intervals
cox_tidy <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)
cox_tidy

# Forest plot of hazard ratios
ggplot(cox_tidy, aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_point() +
  geom_errorbarh(height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(title = "Hazard Ratios with 95% Confidence Intervals",
       x = "Hazard Ratio (log scale)",
       y = "") +
  theme_minimal()
```

# SAS vs R: Statistical Analysis Comparison

Let's summarize the key differences and similarities between SAS and R for statistical analysis:

| Analysis | SAS | R |
|----------|-----|---|
| Descriptive statistics | PROC MEANS, PROC UNIVARIATE | summary(), mean(), sd(), etc. |
| t-test | PROC TTEST | t.test() |
| Chi-square test | PROC FREQ with CHISQ | chisq.test() |
| ANOVA | PROC GLM, PROC ANOVA | aov(), lm() |
| Linear regression | PROC REG, PROC GLM | lm() |
| Logistic regression | PROC LOGISTIC | glm() with family = binomial |
| Survival analysis | PROC LIFETEST, PROC PHREG | survfit(), coxph() |

Key differences:
1. **Output format**: SAS produces formatted tables, while R returns objects that can be further manipulated.
2. **Interactivity**: R allows for more interactive exploration of results.
3. **Visualization**: R makes it easier to create customized visualizations of results.
4. **Model objects**: In R, model results are stored as objects that can be examined, extracted, and used in further analyses.
5. **Default output**: SAS often provides more comprehensive output by default, while R requires explicit requests for specific information.

# Exercises

## Exercise 1: t-test and Visualization

Use the `t.test()` function on the `BP_Change` variable comparing "Drug A" and "Drug B" treatment groups. Interpret the output: what is the p-value and confidence interval? Is the difference statistically significant? Create a boxplot to visualize the difference.

```{r}
# Your code here
```

*Your interpretation here*

## Exercise 2: ANOVA and Linear Regression

Perform a one-way ANOVA to compare `Cholesterol` across the three treatment groups. Check the ANOVA table and note the F-statistic and p-value. Then perform a linear regression with `Cholesterol` as the outcome and `Age`, `Sex`, and `Treatment` as predictors. Identify the coefficients and their p-values, and write a short interpretation.

```{r}
# Your code here
```

*Your interpretation here*

## Exercise 3: Logistic Regression

Fit a logistic regression model with `Outcome` as the dependent variable and `Treatment`, `Age`, and `Baseline_BP` as predictors. Show the summary and compute odds ratios. Determine if treatment group has a significant effect on the odds of success.

```{r}
# Your code here
```

*Your interpretation here*

# Conclusion

In this lesson, you've learned how to perform statistical analysis in R, from basic descriptive statistics to more complex modeling techniques like linear regression, logistic regression, and survival analysis. You've seen how these approaches compare to similar procedures in SAS, and how to interpret the results.

Key takeaways:
- R provides functions for all common statistical analyses that you might perform in SAS
- The formula interface (outcome ~ predictors) is used consistently across different types of models
- Model results in R are stored as objects that can be examined and manipulated
- R makes it easy to visualize data and model results
- The statistical concepts and interpretations are the same, even though the syntax differs

In the next lesson, we'll explore advanced topics including mixed models, Bayesian analysis, high-performance computing, and creating interactive applications with Shiny.

# Session Information

```{r}
sessionInfo()
``` 