---
title: "Lesson 6: Writing Custom Functions and Control Structures in R"
author: "For SAS Users Learning R"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Install required packages if not already installed
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")

# Load the packages
library(tidyverse)  # Loads dplyr and other tidyverse packages
```

# Introduction

Welcome to Lesson 6 of our curriculum for SAS users learning R! In this lesson, we'll explore how to write custom functions and use control structures in R. As a SAS user, you're familiar with macros for code reuse and DATA step logic for control flow. R offers similar capabilities with a more functional programming approach.

## Learning Objectives

By the end of this lesson, you will be able to:

1. Create custom functions in R to automate repetitive tasks
2. Understand the differences between R functions and SAS macros
3. Use control structures like if/else statements and loops
4. Apply functions to data frames using vectorized operations
5. Debug and troubleshoot functions
6. Recognize when to use vectorized operations versus explicit loops

# Functions in R vs Macros in SAS

In SAS, you use macros to generate code and automate repetitive tasks. In R, you use functions. While they serve similar purposes, they work quite differently:

- **SAS macros** generate code that is then executed. They're primarily a text substitution and code generation tool.
- **R functions** are objects that take inputs, perform operations, and return outputs directly. They're first-class objects that can be passed to other functions.

Let's start by creating a simple function in R:

```{r}
# A simple function to calculate BMI
calc_bmi <- function(weight_kg, height_m) {
  bmi <- weight_kg / (height_m^2)
  return(bmi)
}

# Test the function
calc_bmi(70, 1.75)
```
> **Define and Test!** Run the code chunk above. First, R defines the `calc_bmi` function object. Then, the line `calc_bmi(70, 1.75)` calls the function with specific inputs and prints the result.


The structure of an R function is:
1. Name assignment using `<-`
2. The `function()` keyword with parameters in parentheses
3. The function body in curly braces `{}`
4. A `return()` statement (optional, as R returns the last evaluated expression by default)

## Default Arguments and Documentation

Functions can have default arguments and documentation:

```{r}
#' Calculate Body Mass Index (BMI)
#' 
#' @param weight_kg Weight in kilograms
#' @param height_m Height in meters
#' @param round_digits Number of decimal places to round the result
#' @return BMI value (kg/m^2)
#' @examples
#' calc_bmi_improved(70, 1.75)
#' calc_bmi_improved(70, 1.75, round_digits = 1)
calc_bmi_improved <- function(weight_kg, height_m, round_digits = 2) {
  bmi <- weight_kg / (height_m^2)
  bmi <- round(bmi, digits = round_digits)
  return(bmi)
}

# Test with default rounding
calc_bmi_improved(70, 1.75)

# Test with custom rounding
calc_bmi_improved(70, 1.75, round_digits = 1)
```
> **Improved Function:** Run this chunk. Notice how we added a default value for `round_digits`. Try calling it with and without specifying `round_digits`.


The comments starting with `#'` are special Roxygen2 comments that can be used to generate documentation for packages. Even if you're not creating a package, this style of documentation is helpful for understanding what the function does.

## Comparing to SAS Macros

Let's compare our R function to an equivalent SAS macro:

**SAS Macro:**
```
%macro calc_bmi(weight_kg, height_m);
  %let bmi = %sysevalf(&weight_kg / (&height_m ** 2));
  &bmi
%mend calc_bmi;

%put BMI: %calc_bmi(70, 1.75);
```

**R Function:**
```{r}
calc_bmi <- function(weight_kg, height_m) {
  bmi <- weight_kg / (height_m^2)
  return(bmi)
}

cat("BMI:", calc_bmi(70, 1.75), "\n")
```
> **R vs SAS:** Run the R function chunk. Compare the syntax and output to the SAS macro example above it. R functions directly compute and return values.


Key differences:
1. SAS macros use `%` symbols and macro variables with `&`
2. R functions use standard variable assignment
3. SAS macros generate code, while R functions execute code directly
4. R functions are more like SAS PROC or DATA step functions than macros

# Applying Functions to Data

Let's create a sample dataset and apply our BMI function:

```{r}
# Create a sample dataset
patients <- data.frame(
  ID = 1:5,
  Weight_kg = c(70, 65, 80, 85, 60),
  Height_m = c(1.75, 1.60, 1.80, 1.90, 1.65)
)

# Apply the function to calculate BMI for each patient
patients$BMI <- calc_bmi(patients$Weight_kg, patients$Height_m)
> **Vector Power!** Run this chunk (it creates the `patients` data frame first). Notice how `calc_bmi(patients$Weight_kg, patients$Height_m)` works even though the inputs are vectors, not single numbers? R automatically applies the calculation element-wise. The result is assigned to the new `BMI` column.


# View the result
patients
```

Notice how the function automatically works on vectors! This is because R functions are vectorized by default, unlike SAS macros which typically operate on one value at a time.

## Using Functions with dplyr

We can also use our function within dplyr's `mutate()`:

```{r}
# Using the function with mutate()
patients %>%
  mutate(BMI = calc_bmi(Weight_kg, Height_m),
         BMI_Category = case_when(
           BMI < 18.5 ~ "Underweight",
           BMI < 25 ~ "Normal",
           BMI < 30 ~ "Overweight",
> **dplyr Integration:** Run this chunk. You can seamlessly use your custom functions like `calc_bmi` inside `dplyr` verbs like `mutate`. This makes your data wrangling pipelines even more powerful!

           TRUE ~ "Obese"
         ))
```

# Control Structures in R

## Conditional Statements (if/else)

In SAS, you use IF-THEN-ELSE statements in DATA steps. In R, you use `if`, `else if`, and `else`:

```{r}
# Function to categorize BMI
categorize_bmi <- function(bmi) {
  if (bmi < 18.5) {
    return("Underweight")
  } else if (bmi < 25) {
    return("Normal")
  } else if (bmi < 30) {
    return("Overweight")
  } else {
    return("Obese")
> **Conditional Logic:** Run this chunk. Define the `categorize_bmi` function, then test it with different inputs to see the `if/else if/else` logic work.

  }
}

# Test the function
categorize_bmi(22.5)
categorize_bmi(32.1)
```

## Vectorized Conditionals

The `categorize_bmi()` function above works for a single value, but not for vectors. For vectorized operations, use `ifelse()` or `case_when()`:

```{r}
# Vectorized version using case_when
categorize_bmi_vec <- function(bmi) {
  case_when(
    bmi < 18.5 ~ "Underweight",
    bmi < 25 ~ "Normal",
    bmi < 30 ~ "Overweight",
    TRUE ~ "Obese"
  )
}

# Test on a vector
bmi_values <- c(17.5, 22.0, 26.5, 31.0)
> **Vectorized Conditionals:** Run this chunk. The first function (`categorize_bmi`) only works on single values. The second (`categorize_bmi_vec`) uses `case_when` and works on the entire `bmi_values` vector at once. See how it's applied to the `patients` data frame?

categorize_bmi_vec(bmi_values)

# Apply to our dataset
patients %>%
  mutate(BMI_Category = categorize_bmi_vec(BMI))
```

## Loops in R

In SAS, looping is implicit in DATA steps (which process one observation at a time) or explicit with DO loops. In R, you use `for`, `while`, or `repeat` loops:

```{r}
# For loop example
for (i in 1:5) {
  cat("Patient", i, "has BMI:", patients$BMI[i], "\n")
}

> **Loops:** Run this chunk. Observe how the `for` loop iterates through numbers 1 to 5, and the `while` loop continues as long as `i <= 3`. The `cat()` function prints the output for each iteration.

# While loop example
i <- 1
while (i <= 3) {
  cat("Patient", i, "has BMI:", patients$BMI[i], "\n")
  i <- i + 1
}
```

## Vectorization vs. Loops

In R, vectorized operations are generally preferred over loops for performance reasons. Let's compare:

```{r}
# Create a larger dataset
set.seed(123)
large_patients <- data.frame(
  ID = 1:1000,
  Weight_kg = rnorm(1000, mean = 70, sd = 10),
  Height_m = rnorm(1000, mean = 1.7, sd = 0.1)
)

> **Speed Test!** Run this chunk (it creates a larger dataset first). Compare the `system.time` output for the loop versus the vectorized approach. The difference is usually dramatic, highlighting why vectorization is preferred in R!

# Using a loop (slower)
system.time({
  bmi_loop <- numeric(1000)
  for (i in 1:1000) {
    bmi_loop[i] <- calc_bmi(large_patients$Weight_kg[i], large_patients$Height_m[i])
  }
})

# Using vectorization (faster)
system.time({
  bmi_vec <- calc_bmi(large_patients$Weight_kg, large_patients$Height_m)
})
```

As you can see, vectorized operations are much faster in R. This is different from SAS, where loops are often the primary way to process data.

# Apply Functions and Functional Programming

> **Apply to Lists:** Run this chunk. `lapply` applies `mean` to each list element and returns a list. `sapply` does the same but tries to simplify the result into a vector.

R has a family of "apply" functions that apply a function to elements of an object. These are similar to BY-group processing in SAS but more flexible:

## lapply() and sapply()

```{r}
# List of vectors
weight_lists <- list(
  group_a = c(70, 65, 80),
  group_b = c(85, 60, 75)
)

# Apply mean() to each element of the list
lapply(weight_lists, mean)  # Returns a list
sapply(weight_lists, mean)  # Returns a vector or matrix
```

## Using apply() for matrices

```{r}
# Create a matrix of measurements
measurements <- matrix(c(
  # Weight, Height for 5 patients
> **Apply to Matrices:** Run this chunk. `apply` with `MARGIN = 2` applies the function (`mean`) to each column. With `MARGIN = 1`, it applies the custom BMI calculation function to each row.

  70, 1.75,
  65, 1.60,
  80, 1.80,
  85, 1.90,
  60, 1.65
), ncol = 2, byrow = TRUE)

colnames(measurements) <- c("Weight_kg", "Height_m")

# Calculate mean of each column
apply(measurements, 2, mean)  # 2 indicates columns

# Calculate mean of each row
apply(measurements, 1, function(x) x[1] / (x[2]^2))  # Calculate BMI for each row
```

## Using purrr for more complex operations
> **Purrr Power:** Run this chunk. `map_dbl` is like `sapply` but guarantees a numeric vector output. `map2_dbl` applies a function (`~ .x / (.y^2)`) using corresponding elements from two input vectors.


The purrr package (part of tidyverse) provides more consistent and powerful functional programming tools:

```{r}
# Using map functions from purrr
weight_lists %>%
  map_dbl(mean)  # map_dbl ensures a numeric vector is returned

# Map over multiple inputs
map2_dbl(
  large_patients$Weight_kg[1:5],
  large_patients$Height_m[1:5],
  ~ .x / (.y^2)  # Formula notation for the function
)
```

# Real-world Example: Group-wise Statistics Function

Let's create a more complex function that calculates summary statistics by group, similar to what you might do with PROC MEANS in SAS:

```{r}
# Create a function to summarize a numeric variable by group
summarize_by_group <- function(data, group_var, outcome_var) {
  # Convert strings to symbols for dplyr
  group_var <- sym(group_var)
  outcome_var <- sym(outcome_var)
  
  # Use enquo and !! to properly quote the variables
  result <- data %>%
    group_by(!!group_var) %>%
    summarize(
      n = n(),
      mean = mean(!!outcome_var, na.rm = TRUE),
      sd = sd(!!outcome_var, na.rm = TRUE),
      min = min(!!outcome_var, na.rm = TRUE),
      q1 = quantile(!!outcome_var, 0.25, na.rm = TRUE),
      median = median(!!outcome_var, na.rm = TRUE),
      q3 = quantile(!!outcome_var, 0.75, na.rm = TRUE),
      max = max(!!outcome_var, na.rm = TRUE)
    )
  
  return(result)
}
> **Real-world Function:** Run this chunk (it defines the function and creates sample data). This `summarize_by_group` function is quite useful! Test it by summarizing BP by Treatment and Age by Sex. Notice how it uses `sym()` and `!!` (bang-bang) to handle variable names passed as strings – this is a common pattern when writing functions that work with `dplyr`.


# Create a sample clinical dataset
set.seed(123)
clinical_data <- data.frame(
  PatientID = 1:100,
  Treatment = sample(c("Placebo", "Drug A", "Drug B"), 100, replace = TRUE),
  Sex = sample(c("Male", "Female"), 100, replace = TRUE),
  Age = rnorm(100, mean = 50, sd = 10),
  BP = rnorm(100, mean = 120, sd = 15)
)

# Use our function
summarize_by_group(clinical_data, "Treatment", "BP")
summarize_by_group(clinical_data, "Sex", "Age")
```

This function is much more powerful than a SAS macro because:
1. It works with any data frame
2. It can summarize any numeric variable by any grouping variable
3. It returns a data frame that can be further processed

# Error Handling and Debugging

In SAS, you might use `%put` statements or the log to debug macros. In R, you have several options:

> **Error Handling:** Run this chunk. The `safe_calc_bmi` function includes checks using `if` and `stop()`. The `try()` function allows the code to continue even if an error occurs, printing the error message.

## Basic Error Handling

```{r, error=TRUE}
# Function with error handling
safe_calc_bmi <- function(weight_kg, height_m) {
  # Check inputs
  if (!is.numeric(weight_kg) || !is.numeric(height_m)) {
    stop("Weight and height must be numeric")
  }
  
  if (any(height_m <= 0, na.rm = TRUE)) {
    stop("Height must be positive")
  }
  
  # Calculate BMI
  bmi <- weight_kg / (height_m^2)
  return(bmi)
> **Print Debugging:** Run this chunk. Adding `print()` statements inside your function is a simple way to see the values of variables at different stages.

}

# Test with invalid inputs
try(safe_calc_bmi("seventy", 1.75))
try(safe_calc_bmi(70, 0))
```

## Using print() for Debugging

You can add `print()` statements to see intermediate values:

```{r}
debug_calc_bmi <- function(weight_kg, height_m) {
  print(paste("Input weight:", weight_kg))
  print(paste("Input height:", height_m))
  
  bmi <- weight_kg / (height_m^2)
  print(paste("Calculated BMI:", bmi))
  
  return(bmi)
}

debug_calc_bmi(70, 1.75)
```

## Using RStudio's Debugging Tools

RStudio provides interactive debugging tools:
- `browser()`: Pauses execution and allows interactive inspection
- `debug()`: Flags a function for step-by-step debugging
- `traceback()`: Shows the call stack after an error

These are more advanced than what's available in SAS and make debugging complex functions easier.

# SAS vs R: Function Programming Comparison

Let's summarize the key differences between SAS macros and R functions:

| Aspect | SAS Macros | R Functions |
|--------|------------|-------------|
| Purpose | Code generation | Direct computation |
| Syntax | `%macro`, `%mend`, `&` for variables | `function()`, `{}`, standard variable names |
| Scope | Global by default | Local by default |
| Return values | Text substitution | Any R object |
| Vectorization | No built-in vectorization | Naturally vectorized |
| Debugging | Log messages | Interactive debugging tools |
| Reusability | Can be stored in macro libraries | Can be stored in packages or scripts |
| First-class objects | No | Yes (can be passed to other functions) |

For SAS users, the key mindset shift is understanding that:
1. R functions operate on entire vectors or data frames at once
2. R functions return objects that can be used in further calculations
3. R's functional programming approach is more flexible than SAS's macro language

# Exercises

## Exercise 1: Creating a BMI Function

Write a custom function `calc_bmi(weight_kg, height_m)` that calculates body mass index (BMI = weight (kg) / height (m)^2) and returns the BMI value. Test your function on a few sample inputs. Then use your function within a `mutate()` call to add a BMI column to a patient data frame (with weight and height columns). **Type your code in the chunk below and run it.**

```{r}
# Your code here
```

## Exercise 2: Group Summary Function

Create a function `summarize_by_group(data, group_var, outcome_var)` that takes a data frame and two column names (group and outcome) and returns a data frame with count, mean, and standard deviation of the outcome for each group. Test this function on a dataset (e.g., summarize blood pressure by treatment group). **Type your code in the chunk below and run it.**

```{r}
# Your code here
```

## Exercise 3: Loops and Conditionals

Write a loop that goes through numbers 1 to 10, and prints whether each number is even or odd. Then, consider how you might achieve the same result without an explicit loop (hint: R can do `ifelse()` on a vector directly). **Type your code in the chunk below and run it.**

```{r}
# Your code here
```

# Conclusion

In this lesson, you've learned how to write custom functions and use control structures in R. You've seen how R's functional programming approach differs from SAS's macro language, and how to leverage vectorization for efficient code.

Key takeaways:
- R functions are more powerful and flexible than SAS macros
- Vectorized operations are preferred over loops in R for performance
- Control structures like if/else and for loops work similarly to SAS but with different syntax
- The apply family of functions and purrr provide powerful tools for functional programming
- Error handling and debugging in R offer more interactive options than SAS

In the next lesson, we'll explore statistical analysis in R, including hypothesis tests, regression models, and survival analysis.

# Session Information

```{r}
sessionInfo()
``` 